---
title: 'FDA Drug Trials Snapshots 2015-2019 <br> Race, ethnicty participation in clinical trials'
author: "Ariel Carmeli"
date: "December 2020"
output:
  html_document:
    toc: yes
    toc_depth: 2
  pdf_document:
    toc: yes
    toc_depth: '2'
---

```{r setup, include=FALSE}
library(tidyverse)
library(forcats)
library(knitr)
```


```{r, include=FALSE}
# Read in clinical trial data
fda_approvals <- read.csv('FDA_Drug_Trials_Snapshots_2015-19.csv')

# Change class type of select variables to aid in data processing and visualization
fda_approvals$Trial_size <- as.numeric(as.character(fda_approvals$Trial_size))
fda_approvals$Therapeutic_Area <- as.character(fda_approvals$Therapeutic_Area)
fda_approvals$Brand_Name <- as.character(fda_approvals$Brand_Name)
fda_approvals$United_States <- as.numeric(as.character(fda_approvals$United_States))

# Add a column to group trial size
fda_approvals <- fda_approvals %>% 
  mutate( Trial_size_grouped = case_when(
    Trial_size < 100 ~ "a. 1-99 patients", 
    Trial_size >= 100 & Trial_size < 200 ~ "b. 100-199 patients", 
    Trial_size >= 200 & Trial_size < 300 ~ "c. 200-299 patients",
    Trial_size >= 300 & Trial_size < 500 ~ "d. 300-499 patients",
    Trial_size >= 500 & Trial_size < 1000 ~ "e. 500-999 patients",
    Trial_size >= 1000 & Trial_size < 2000 ~ "f. 1000-1999 patients",
    Trial_size >= 2000  ~ "g. Over 2000 patients"
  ), .after = Trial_size)

# Add a column for non-hispanic 
fda_approvals <- fda_approvals %>% mutate(Non_Hispanic = 100 - Hispanic, .after = Hispanic)

# Create longer version, for plotting
fda_approvals_long <- pivot_longer(fda_approvals, cols = Women:Age_80_or_older, names_to = "Demographic", values_to = "Percentage")

fda_approvals_long <- fda_approvals_long %>% pivot_longer(cols = Sex_comparison:Age_comparison, names_to = "Comparison", values_to = "Compared")

# Change class type of variable in long
fda_approvals_long$Percentage <- as.numeric(as.character(fda_approvals_long$Percentage))

```


```{r}
# Missing data
no_approvals <- fda_approvals %>% 
  group_by(Approval_Year) %>% 
  summarize(Approvals = n()) 

reported_values <- fda_approvals_long %>% 
  filter(Percentage != "NA") %>% 
  filter(Demographic == "White" | Demographic == "Black" | Demographic == "Asian" | Demographic == "Hispanic") %>% 
  group_by(Approval_Year, Demographic) %>% 
  summarize(Non_NR = n() / 3) %>% 
  pivot_wider(names_from = Demographic, values_from = Non_NR, values_fill = 0)

missing_data <- data.frame(no_approvals$Approval_Year, 
                           no_approvals$Approvals, 
                           reported_values$Asian,
                           reported_values$Black,
                           reported_values$White,
                           reported_values$Hispanic)

names(missing_data) <- c("Approval_Year", "Approvals", "Asian", "Black", "White", "Hispanic")

missing_data <- missing_data %>% 
  mutate(Asian = round(100 * Asian / Approvals, 0),
         Black = round(100 * Black / Approvals, 0),
         White = round(100 * White / Approvals, 0),
         Hispanic = round(100 * Hispanic / Approvals, 0))

```

```{r}

```



# 0) Demonstrate data comparability with [2015-2019 FDA Drug Trials Snapshots Summary Report](https://www.fda.gov/media/143592/download)

**Total number of trial participants (Global)**

*Page 4 in FDA Summary*
```{r}

total_patients <- sum(fda_approvals$Trial_size)

print(paste("The total number of trial participants in our dataset is:", total_patients))
print(paste("The total number of trial participants in FDA report is: 292537"))
print(paste("Our dataset has", 292537- sum(fda_approvals$Trial_size), "fewer participants, which is within", 
  round(100*((292537- sum(fda_approvals$Trial_size))/292537), 1), "percent of the FDA total"))

```

**Race distributions of trial participants (Global)**

*Page 9 in FDA Summary*
```{r, include = FALSE}

# Find number of participants by race for each approval  
global_black_participation <- fda_approvals %>% 
  select(Brand_Name, Trial_size, Black) %>% 
  filter(Black != "NA") %>% 
  mutate(Black_participants = (Black/100) * Trial_size)

global_white_participation <- fda_approvals %>% 
  select(Brand_Name, Trial_size, White) %>% 
  filter(White != "NA") %>% 
  mutate(White_participants = (White/100) * Trial_size)

global_asian_participation <- fda_approvals %>% 
  select(Brand_Name, Trial_size, Asian) %>% 
  filter(Asian != "NA") %>% 
  mutate(Asian_participants = (Asian/100) * Trial_size)

demographic_participation <- c(round(sum(100 * global_asian_participation$Asian_participants) /
                                       sum(global_asian_participation$Trial_size), 0), 
                               round(sum(100 * global_black_participation$Black_participants) / 
                                       sum(global_black_participation$Trial_size), 0),
                               round(sum(100 * global_white_participation$White_participants) / 
                                       sum(global_white_participation$Trial_size), 0))

fda_snapshots <- c(11, 7, 76)
races <- c("Asian", "Black or African American", "White")
  
Race_distribution_comparison <- data.frame(races, fda_snapshots, demographic_participation)
names(Race_distribution_comparison) <- c("Race", "FDA_Snapshots", "Our_Database")

Race_distribution_comparison_graph <- Race_distribution_comparison %>% 
  pivot_longer(cols = FDA_Snapshots:Our_Database, names_to = "Database", values_to = "Participation") %>% 
  ggplot(aes(Race, Participation, fill = Database)) +
  geom_bar(stat = "identity", position = 'dodge') +
  geom_text(aes(label=Participation), position = position_dodge(0.9), vjust=-0.3, size=3) +
  xlab("Race") +
  ylab("Participation") + 
  #theme(text = element_text(size=15)) +
  ggtitle("Race distribution of trial participants")

```

```{r}
Race_distribution_comparison_graph
```


```{r, fig.width = 5}
# Find number of participants by hispanic ethnicity for each approval  

global_hispanic_participation <- fda_approvals %>% 
  select(Brand_Name, Trial_size, Hispanic) %>% 
  filter(Hispanic != "NA") %>% 
  mutate(Hispanic_participants = (Hispanic/100) * Trial_size)

hispanic_participation_rate <- round(sum(100 * global_hispanic_participation$Hispanic_participants) /
                                       sum(global_hispanic_participation$Trial_size), 0)

hispanic_participation_rate

# Plot actual vs. expected participation

calculated_demographic_participation <- c(round(sum(100 * global_asian_participation$Asian_participants) /
                                       sum(global_asian_participation$Trial_size), 1), 
                               round(sum(100 * global_black_participation$Black_participants) / 
                                       sum(global_black_participation$Trial_size), 1),
                               round(sum(100 * global_white_participation$White_participants) / 
                                       sum(global_white_participation$Trial_size), 2), 
                               round(sum(100 * global_hispanic_participation$Hispanic_participants) /
                                       sum(global_hispanic_participation$Trial_size), 1))

expected_participation <- c(6.8, 14.3, 76.5, 18.5)
demographics <- c("Asian", "Black or African American", "White", "Hispanic")
  
actual_vs_expected_participation <- data.frame(demographics, expected_participation,
                                               calculated_demographic_participation)

names(actual_vs_expected_participation) <- c("Demographic", "Expected", "Actual")

actual_vs_expected_participation

fda_approvals$Trial_size <- as.numeric(as.character(fda_approvals$Trial_size))

actual_vs_expected_participation %>% 
  pivot_longer(cols = Expected:Actual, names_to = "Label", values_to = "Participation") %>% 
  ggplot(aes(Demographic, Participation, fill = Label)) +
  #geom_point() +
  geom_bar(stat = "identity", position = 'dodge') +
  geom_text(aes(label=Participation), position = position_dodge(0.9), vjust = -0.4, size=4) +
  theme(legend.position = "top", legend.title = element_blank()) +
  xlab("Demographic") +
  ylab("Average participation") + 
  #theme(text = element_text(size=15)) +
  ggtitle("Average trial participation by race, ethnicity (2015-19)")

```


**Clinical Trial Participation by Therapeutic Area**

*Page 30 in FDA Summary*

```{r, include=FALSE}

fda_approvals_check <- fda_approvals %>% 
  select(Therapeutic_Area, Trial_size) %>% 
  mutate(Therapeutic_Area = case_when(
    Therapeutic_Area == "Oncology" ~ "Oncology and Hematology", 
    Therapeutic_Area == "Hematology" ~ "Oncology and Hematology",
    Therapeutic_Area == "Hematology (Sickle cell)" ~ "Oncology and Hematology",
    Therapeutic_Area == "Endocrinology and Metabolism" ~ "Endocrinology and Metabolism",
    Therapeutic_Area == "Infectious Diseases" ~ "Infectious Diseases",
    Therapeutic_Area == "Neurology" ~ "Neurology",
    Therapeutic_Area == "Gynecology" ~ "Gynecology",
    Therapeutic_Area == "Dermatology" ~ "Dermatology",
    Therapeutic_Area == "Pulmonology and Rheumatology" ~ "Pulmonology and Rheumatology",
    Therapeutic_Area == "Gastroenterology" ~ "Gastroenterology",
    Therapeutic_Area == "Psychiatry" ~ "Psychiatry",
    Therapeutic_Area == "Sleep Disorders" ~ "Psychiatry",
    Therapeutic_Area == "Ophthalmology" ~ "Ophthalmology",
    Therapeutic_Area == "Anesthesia and Analgesia" ~ "Anesthesia and Analgesia",
    Therapeutic_Area == "Medical Imaging" ~ "Medical Imaging",
    Therapeutic_Area == "Cardiovascular Diseases" ~ "Cardiovascular Diseases"
  )) %>% 
  group_by(Therapeutic_Area) %>% 
  summarize(Our_database = sum(Trial_size)) %>% 
  arrange(desc(Our_database))

# Cardio, Onc, Endocrin, ID, Neurology, Derm, Gyn, Pulm/Rheum, GI, Psychiatry, Anaesthesia, Ophthal, Medical Imaging
fda_snapshots <- c(59000, 35000, 41000, 32500, 26000, 20500, 21000, 20000, 15000, 10500, 6500, 4500, 1500)

Total_patients_check <- data.frame(fda_approvals_check$Therapeutic_Area, 
                                      fda_approvals_check$Our_database, 
                                      fda_snapshots)
names(Total_patients_check) <- c("Therapeutic_Area", "Our_Database", 
                                    "FDA_Snapshots")

fda_approvals_check_long <- pivot_longer(Total_patients_check, cols = Our_Database:FDA_Snapshots, 
                                         names_to = "Database", values_to = "Participants")

Patient_participation_graph <- fda_approvals_check_long %>% 
  ggplot(aes(reorder(Therapeutic_Area, Participants), Participants, fill=Database)) +
  geom_bar(stat = "identity", position = 'dodge') +
  geom_text(aes(label=Participants), position = position_dodge(0.9), hjust=-0.15, size=5) +
  coord_flip() + 
  xlab("Therapeutic Area") +
  ylab("Number of Participants") + 
  theme(text = element_text(size=20)) +
  ggtitle("Patient participation by Therapeutic Area [FDA approvals 2015-19]")

```

```{r, fig.height = 10, fig.width = 20}
Patient_participation_graph
```

# 1) Summary Descriptive Statistics

**a) Number of Approvals by Year**
```{r, include = FALSE}
approvals_table_year <- fda_approvals %>% 
  group_by(Approval_Year) %>% 
  summarise(Number_of_approvals = n())
```

```{r}
approvals_table_year
```

**b) Approvals and Patient Participation by Therapeutic Area**
```{r, include = FALSE}

approvals_table_TA <- fda_approvals %>% 
  select(Therapeutic_Area, Trial_size) %>% 
  filter(Trial_size != "NA") %>% 
  group_by(Therapeutic_Area) %>% 
  summarise(Total_patients = sum(Trial_size), Number_of_approvals = n()) %>% 
  mutate(Avg_patients_per_approval = Total_patients/Number_of_approvals) %>% 
  mutate(Avg_patients_per_approval = round(Avg_patients_per_approval, 0)) %>% 
  arrange(desc(Total_patients))

```

```{r, fig.height = 5, fig.width = 15}
approvals_table_TA
```


```{r, include = FALSE}
# Plot trial size by Therapeutic Area

approvals_box <- fda_approvals %>% 
  select(Therapeutic_Area, Trial_size) %>% 
  filter(Trial_size != "NA") %>% 
  mutate(Therapeutic_Area = fct_reorder(Therapeutic_Area, Trial_size, .fun='median')) %>%
  ggplot(aes(Therapeutic_Area, Trial_size)) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle=90)) +
  theme(text = element_text(size=12)) +
  coord_flip() +
  xlab("Therapeutic Area") +
  ylab("Participants") +
  ggtitle("Participants per approval")

approvals_box_jitter <- fda_approvals %>% 
  select(Therapeutic_Area, Trial_size) %>% 
  filter(Trial_size != "NA") %>% 
  mutate(Therapeutic_Area = fct_reorder(Therapeutic_Area, Trial_size, .fun='median')) %>%
  ggplot(aes(Therapeutic_Area, Trial_size)) +
  geom_boxplot() +
  geom_jitter() +
  theme(axis.text.x = element_text(angle=90)) +
  theme(text = element_text(size=12)) +
  coord_flip() +
  xlab("Therapeutic Area") +
  ylab("Participants") +
  ggtitle("Participants per approval")

```

```{r, fig.height = 5, fig.width = 15}

approvals_box
approvals_box_jitter

```


**c) Median representation of race and ethnicity by trial size**

*For comparison, Getz et. al. (2020) uses an expected distribution of race/ethnicity in the U.S. population*

- Asian: 3.5%

- Black: 15.6%

- Hispanic: 8.2%

- White: 67.1%


```{r, include = FALSE}
# Create a helper function to summarize median demographic

median_demographics <- function(fda_approvals){
  
  Asian <- fda_approvals %>% 
    select(Brand_Name, Trial_size_grouped, Asian, United_States) %>% 
    group_by(Trial_size_grouped) %>% 
    filter(Asian != "NA") %>% # Filter out "not reported"
    filter(Trial_size_grouped != "NA") %>% # Filter out "Unknown" trial size
    #filter(United_States >= US_proportion) %>% 
    summarise(Median_Asian = median(Asian), Number_of_trials = n()) %>% 
    arrange(Trial_size_grouped)
  
  Black <- fda_approvals %>% 
    select(Brand_Name, Trial_size_grouped, Black, United_States) %>% 
    group_by(Trial_size_grouped) %>% 
    filter(Black != "NA") %>%
    filter(Trial_size_grouped != "NA") %>%
    #filter(United_States >= US_proportion) %>% 
    summarise(Median_Black = median(Black), Number_of_trials = n()) %>% 
    arrange(Trial_size_grouped)
  
  Hispanic <- fda_approvals %>% 
    select(Brand_Name, Trial_size_grouped, Hispanic, United_States) %>% 
    group_by(Trial_size_grouped) %>% 
    filter(Hispanic != "NA") %>%
    filter(Trial_size_grouped != "NA") %>%
    #filter(United_States >= US_proportion) %>% 
    summarise(Median_Hispanic = median(Hispanic), Number_of_trials = n()) %>% 
    arrange(Trial_size_grouped)
    
  White <- fda_approvals %>% 
    select(Brand_Name, Trial_size_grouped, White, United_States) %>% 
    group_by(Trial_size_grouped) %>% 
    filter(White != "NA") %>%
    filter(Trial_size_grouped != "NA") %>%
    #filter(United_States >= US_proportion) %>% 
    summarise(Median_White = median(White), Number_of_trials = n()) %>% 
    arrange(Trial_size_grouped)
  
  No_trials <- fda_approvals %>% 
    group_by(Trial_size_grouped) %>% 
    filter(Trial_size_grouped != "NA") %>%
    #filter(United_States >= US_proportion) %>% 
    summarise(Number_of_trials = n())
  
  Race_size_description <- data.frame(No_trials$Trial_size_grouped, 
                                      No_trials$Number_of_trials, 
                                      Asian$Median_Asian, 
                                      Black$Median_Black, 
                                      Hispanic$Median_Hispanic, 
                                      White$Median_White
                                      )
  names(Race_size_description) <- c("Trial size", "Number of trials", 
                                    "Median Asian", "Median Black", "Median Hispanic", 
                                    "Median White")
    
  return(Race_size_description)
}
```


**All Clinical Trials**

```{r}

Race_size_description <- median_demographics(fda_approvals)
kable(Race_size_description, format = "html")

```

**Clinical Trials with >= 50% U.S. representation**

```{r}

fda_approvals_us_prop <- fda_approvals %>% filter(United_States >= 50)

Race_size_description_US_majority <- median_demographics(fda_approvals_us_prop)
kable(Race_size_description_US_majority, format = "html")

```

```{r, fig.height = 5, fig.width = 15}

test <- Race_size_description %>% 
  pivot_longer(cols = `Median Asian`:`Median White`, names_to = "Demographic", values_to = "Percentage")

test %>% 
  filter(Demographic != "Median White") %>%
  ggplot(aes(`Trial size`, y=Percentage, group=Demographic)) +
  geom_line()
  #geom_bar(stat = "identity", position = 'dodge')
  
```


**d) Cumulative distribution by demographic**

```{r, fig.height = 5, fig.width = 10}
# Race
fda_approvals_long %>% 
  filter(Demographic == "White" | Demographic == "Black" | Demographic == "Asian" | Demographic == "Hispanic") %>% 
  ggplot(aes(Percentage/100, colour = Demographic)) + stat_ecdf(geom = "step") +
  #facet_wrap(~Demographic, ncol=3) + 
  scale_x_continuous(breaks = round(seq(min(0), max(1), by = 0.05),1)) +
  scale_y_continuous(breaks = round(seq(min(0), max(1), by = 0.1),1)) +
  theme(axis.title.x = element_text(size=15), 
        axis.title.y = element_text(size=15),
        axis.text.x = element_text(size=15),
        axis.text.y = element_text(size=15),
        title = element_text(size=15),
        legend.text=element_text(size=20), 
        legend.position = "top", 
        legend.title = element_blank()
        ) +
  ylab("Cumulative percentage of FDA approvals") +
  xlab("Relative participation in clinical trials") +
  ggtitle("Cumulative distribution of race and ethnicity participation in trials (2015-19)")


```

```{r}
# Ethnicity
fda_approvals %>% 
  ggplot(aes(Hispanic)) + stat_ecdf(geom = "step") +
  scale_x_continuous(breaks = round(seq(min(0), max(100), by = 5),1)) +
  scale_y_continuous(breaks = round(seq(min(0), max(1), by = 0.1),1)) +
  theme(text = element_text(size=10)) +
  ggtitle("Cumulative distribution of ethnicity clinical trial participation in FDA approvals [2015-19]")

```

**e) Median representation by Race/TA**


```{r}

# TABLECHECK -- Median representation by race and TA
Black_TA_view <- fda_approvals %>% 
  group_by(Therapeutic_Area, Disease) %>% 
  filter(Black != "NA") %>% 
  summarize(Median_Black = median(Black), Number_of_Approvals = n()) %>% 
  arrange(Therapeutic_Area)

Asian_TA_view <- fda_approvals %>% 
  group_by(Therapeutic_Area, Disease) %>% 
  filter(Asian != "NA") %>% 
  summarize(Median_Asian = median(Asian), Number_of_Approvals = n()) %>% 
  arrange(Therapeutic_Area)

Hispanic_TA_view <- fda_approvals %>% 
  group_by(Therapeutic_Area, Disease) %>% 
  filter(Hispanic != "NA") %>% 
  summarize(Median_Hispanic = median(Hispanic), Number_of_Approvals = n()) %>% 
  arrange(Therapeutic_Area)

White_TA_view <- fda_approvals %>% 
  group_by(Therapeutic_Area, Disease) %>% 
  filter(White != "NA") %>% 
  summarize(Median_White = median(White), Number_of_Approvals = n()) %>% 
  arrange(Therapeutic_Area)

#Other_TA_view <- fda_approvals %>% 
#  group_by(Therapeutic_Area, Disease) %>% 
#  filter(Other != "NA") %>% 
#  summarize(Median_Other = median(Other), Number_of_Approvals = n()) %>% 
#  arrange(Therapeutic_Area)

All_approvals <- fda_approvals %>% 
  group_by(Therapeutic_Area, Disease) %>% 
  summarise(Number_of_approvals = n()) %>% 
  arrange(Therapeutic_Area)

Race_TA_table <- data.frame(All_approvals$Therapeutic_Area,
                            All_approvals$Number_of_approvals,
                            Asian_TA_view$Median_Asian,
                            Black_TA_view$Median_Black,
                            White_TA_view$Median_White)

names(Race_TA_table) <- c("Therapeutic Area", "Number of approvals", "Asian", "Black", "White")

Race_TA_table

```


```{r}

# Count of trials by size

t <- fda_approvals_long %>% 
  filter(Demographic == "Black" | Demographic == "White" | Demographic == "Asian") %>% 
  group_by(Demographic, Percentage) %>% 
  summarize(Count = n()/3, Relative = round(Count/235, 2)) %>% 
  mutate(Participation = case_when(
    Percentage == 0 ~ "a. Zero", 
    Percentage == 1 ~ "b. One", 
    Percentage == 2 ~ "c. Two", 
    Percentage == 3 ~ "d. Three",
    Percentage == 4 ~ "e. Four", 
    Percentage == 5 ~ "f. Five",
    Percentage >= 6 & Percentage <= 10 ~ "g. 6-10",
    Percentage >= 11 & Percentage <= 15 ~ "h. 11-15",
    Percentage >= 16 & Percentage <= 20 ~ "i. 16-20",
    Percentage >= 20 & Percentage <= 70 ~ "j. 20-70",
    Percentage > 70 ~ "k. >70"
    )) %>% 
  group_by(Demographic, Participation) %>% 
  summarize(Count = sum(Count), Relative = sum(Relative))

t %>% filter(Participation != 0) %>% 
  filter(Participation != "j. 20-70") %>% 
  filter(Participation != "k. >70") %>% 
  ggplot(aes(Participation, y=Count, width=1, fill=Demographic)) +
  #geom_col(stat = "identity")
  geom_bar(stat = "identity", position = 'dodge') + 
  geom_text(aes(label=Count), position = position_dodge(0.9), vjust=-0.3, size=4) +
  xlab("Percent participation in trials for FDA approval") +
  ylab("Number of FDA approvals") +
  ggtitle("Number of FDA approvals that had <20% participation in trials by race")
  #facet_wrap(~Demographic)

t2 <- t %>% select(-Relative) %>% pivot_wider(names_from = Demographic, values_from = Count)

t2

#Patient_participation_graph <- fda_approvals_check_long %>% 
#  ggplot(aes(reorder(Therapeutic_Area, Participants), Participants, fill=Database)) +
#  geom_bar(stat = "identity", position = 'dodge') +
#  geom_text(aes(label=Participants), position = position_dodge(0.9), hjust=-0.15, size=5) +
#  coord_flip() + 
#  xlab("Therapeutic Area") +
#  ylab("Number of Participants") + 
#  theme(text = element_text(size=20)) +
#  ggtitle("Patient participation by Therapeutic Area [FDA approvals 2015-19]")

```

```{r, fig.width = 5}
trend <- read.csv('Trend.csv')

trend_long <- trend %>% pivot_longer(cols = Black:Hispanic, names_to = "Demographic", values_to = "Participation")

trend_long %>% 
  filter(Demographic != "Other") %>% 
  #filter(Demographic == "White" | Demographic == "Asian" | Demographic == "Black" ) %>% 
  ggplot(aes(Year, y=Participation, colour = Demographic)) + 
  geom_line() +
  geom_text(aes(label=Participation), vjust = -0.4, size=3) + 
  theme(legend.position = "top", legend.title = element_blank()) +
  ggtitle("Average trial participation over time by race, ethnicity")
  #geom_col() +
  #geom_bar(stat = "identity", position = 'dodge')

```


# 2) Race, Ethnicity, Size, Geography through individual drug approvals

```{r, include = FALSE}
# Race participation 
race_participation_graph <- fda_approvals_long %>% 
  select(Brand_Name, Demographic, Percentage, Approval_Year) %>% 
  filter(Demographic == "White" | Demographic == "Black" | Demographic == "Asian" | Demographic == "Other") %>%
  filter(Percentage > -1) %>% 
  unique() %>% 
  ggplot(aes(Demographic, Percentage)) + 
  geom_jitter(width = 0.2, aes(colour=Demographic)) + 
  scale_y_continuous(breaks = round(seq(min(0), max(100), by = 5),1)) +
  ggtitle("a.i Race participation in clinical trials for FDA approvals [2015-19]")

race_participation_trend_graph <- fda_approvals_long %>% 
  select(Brand_Name, Demographic, Percentage, Approval_Year) %>% 
  filter(Demographic == "White" | Demographic == "Black" | Demographic == "Asian" | Demographic == "Other") %>%
  filter(Percentage > -1) %>% 
  unique() %>% 
  ggplot(aes(Demographic, Percentage)) + 
  geom_jitter(width = 0.2, aes(colour=Approval_Year)) + 
  scale_y_continuous(breaks = round(seq(min(0), max(100), by = 5),1)) +
  ggtitle("a.ii Race participation by Year in clinical trials for FDA approvals [2015-19]")

# Ethnicity participation 

ethnicity_participation_graph <- fda_approvals_long %>% 
  select(Brand_Name, Demographic, Percentage, Approval_Year) %>% 
  filter(Demographic == "Hispanic" | Demographic == "Non_Hispanic") %>%
  filter(Percentage > -1) %>% 
  unique() %>% 
  ggplot(aes(Demographic, Percentage)) + 
  geom_jitter(width = 0.2, aes(colour=Demographic)) + 
  scale_y_continuous(breaks = round(seq(min(0), max(100), by = 5),1)) +
  ggtitle("b.i Ethnicity participation in clinical trials for FDA approvals [2017-19]")

ethnicity_participation_trend_graph <- fda_approvals_long %>% 
  select(Brand_Name, Demographic, Percentage, Approval_Year) %>% 
  filter(Demographic == "Hispanic" | Demographic == "Non_Hispanic") %>%
  filter(Percentage > -1) %>% 
  unique() %>% 
  ggplot(aes(Demographic, Percentage)) + 
  geom_jitter(width = 0.2, aes(colour=Approval_Year)) + 
  scale_y_continuous(breaks = round(seq(min(0), max(100), by = 5),1)) +
  ggtitle("b.ii Ethnicity participation by Year in clinical trials for FDA approvals [2017-19]")

```

```{r, fig.width = 5}
race_participation_graph
race_participation_trend_graph
ethnicity_participation_graph
ethnicity_participation_trend_graph
```

```{r, include = FALSE}
# Race and US participation (all years)
race_geography_graph <- fda_approvals_long %>% 
  select(Brand_Name, Demographic, Percentage, United_States, Therapeutic_Area) %>% 
  filter(Demographic == "White" | Demographic == "Black" | Demographic == "Asian" | Demographic == "Hispanic") %>%
  filter(Percentage != "NA") %>% 
  filter(United_States != "NA") %>% 
  unique() %>% 
  ggplot(aes(United_States, Percentage)) + 
  geom_jitter(width = 0.2) + 
  facet_wrap(~Demographic, ncol = 4) + 
  xlab("US representation") +
  ylab("Demographic representation") +
  ggtitle("Race/Ethnicity and Geography in FDA approvals (2015-19)")

```

```{r, fig.height = 5, fig.width = 12}
race_geography_graph
```

```{r}
fda_approvals_long %>% 
  select(Brand_Name, Approval_Year, Demographic, Percentage, United_States, Therapeutic_Area) %>% 
  filter(Demographic == "White" | Demographic == "Black" | Demographic == "Asian" | Demographic == "Hispanic") %>%
  #filter(Percentage != "NA") %>% 
  #filter(United_States != "NA") %>% 
  unique() %>% 
  ggplot(aes(Approval_Year, Percentage)) + 
  geom_jitter(width = 0.2) + 
  facet_wrap(~Demographic, ncol = 2) + 
  xlab("Approval Year") +
  ylab("Demographic representation") +
  ggtitle("Race/Ethnicity and Geography in FDA approvals (2015-19)")
```


```{r, include = FALSE}
# Ethnicity and US participation (2017-19 only)
ethnicity_geography_graph <- fda_approvals_long %>% 
  select(Brand_Name, Demographic, Percentage, United_States, Therapeutic_Area) %>% 
  #filter(Demographic == "Hispanic" | Demographic == "Non_Hispanic") %>%
  filter(Demographic == "Hispanic") %>%
  filter(Percentage != "NA") %>% 
  filter(United_States != "NA") %>% 
  unique() %>% 
  ggplot(aes(United_States, Percentage)) + 
  geom_jitter(width = 0.2) + 
  facet_wrap(~Demographic) + 
  xlab("US representation") +
  ylab("Ethnicity representation") +
  ggtitle("d. Ethnicity, geography in FDA approvals (2017-19)")

```


```{r, fig.height = 5, fig.width = 3}
ethnicity_geography_graph
```

```{r, include = FALSE}
# Race participation and Trial size (2015-19)
race_size_graph <- fda_approvals_long %>% 
  select(Brand_Name, Demographic, Percentage, Therapeutic_Area, Trial_size, United_States) %>% 
  filter(Demographic == "White" | Demographic == "Black" | Demographic == "Asian") %>%
  filter(Percentage != "NA") %>% 
  unique() %>% 
  ggplot(aes(Trial_size, Percentage)) + 
  geom_jitter(width = 0.2) +
  facet_wrap(~Demographic, ncol = 4) + 
  xlim(0, 3500) +
  #ylim(0, 25) +
  xlab("Number of clinical trial participants") +
  ylab("Race representation") +
  ggtitle("e. Race representation and number of clinical trial participants in FDA approvals [2015-19]")

```

```{r, fig.height = 5, fig.width = 10}
race_size_graph
```

```{r, include = FALSE}
# Ethnicity participation and Trial size (2015-19)
ethnicity_size_graph <- fda_approvals_long %>% 
  select(Brand_Name, Demographic, Percentage, Therapeutic_Area, Trial_size, United_States) %>% 
  filter(Demographic == "Hispanic" | Demographic == "Non_Hispanic") %>%
  filter(Percentage != "NA") %>% 
  unique() %>% 
  ggplot(aes(Trial_size, Percentage)) + 
  geom_jitter(width = 0.2) +
  facet_wrap(~Demographic) + 
  xlim(0, 3500) +
  #ylim(0, 25) +
  xlab("Number of clinical trial participants") +
  ylab("Ethnicity representation") +
  ggtitle("f. Ethnicity representation and number of clinical trial participants in FDA approvals [2017-19]")

```

```{r, fig.height = 5, fig.width = 10}
ethnicity_size_graph
```

# 3) Therapeutic Area deep-dive: Race and Ethnicity representation individual drug approvals

```{r, fig.height = 6, fig.width = 10}
# For White Paper

# Demographic TA graph. Race/Ethnicity together
TA_demo_graph <- fda_approvals_long %>% 
  select(Brand_Name, Demographic, Percentage, Therapeutic_Area, Approval_Year) %>% 
  filter(Demographic == "White" | Demographic == "Black" | Demographic == "Asian" | Demographic == "Hispanic") %>%
  filter(Percentage > -1) %>% 
  unique() %>% 
  ggplot(aes(Demographic, Percentage)) + 
  geom_jitter(width = 0.2, aes(colour=Demographic)) + 
  scale_y_continuous(breaks = round(seq(min(0), max(100), by = 10),1)) +
  facet_wrap(~Therapeutic_Area, ncol = 5) + 
  theme(legend.position = "top", legend.title = element_blank()) +
  ggtitle("Race/Ethnicity participation by Therapeutic Area in clinical trials for FDA approvals [2015-19]")

TA_demo_graph

```


```{r, fig.height = 8, fig.width = 10}
# Oncology deep dive
fda_approvals_long %>% 
  filter(Therapeutic_Area == "Oncology") %>% 
  select(Brand_Name, Demographic, Percentage, Therapeutic_Area, Disease, Approval_Year) %>% 
  filter(Demographic == "White" | Demographic == "Black" | Demographic == "Asian" | Demographic == "Hispanic") %>%
  filter(Percentage > -1) %>% 
  unique() %>% 
  ggplot(aes(Demographic, Percentage)) + 
  geom_jitter(width = 0.2, aes(colour=Demographic)) + 
  scale_y_continuous(breaks = round(seq(min(0), max(100), by = 10),1)) +
  facet_wrap(~Disease) + 
  theme(legend.position = "top", legend.title = element_blank()) +
  ggtitle("Diversity in Oncology FDA approvals [2015-19]")
```

```{r, fig.height = 5, fig.width = 10}

# Infectious Disease deep dive
fda_approvals_long %>% 
  filter(Therapeutic_Area == "Infectious Diseases") %>% 
  select(Brand_Name, Demographic, Percentage, Therapeutic_Area, Disease, Approval_Year) %>% 
  filter(Demographic == "White" | Demographic == "Black" | Demographic == "Asian" | Demographic == "Hispanic") %>%
  filter(Percentage > -1) %>% 
  unique() %>% 
  ggplot(aes(Demographic, Percentage)) + 
  geom_jitter(width = 0.2, aes(colour=Demographic)) + 
  scale_y_continuous(breaks = round(seq(min(0), max(100), by = 10),1)) +
  facet_wrap(~Disease) + 
  theme(legend.position = "top", legend.title = element_blank()) +
  ggtitle("Diversity in Infectious Diseases FDA approvals [2015-19]")

```

```{r, fig.height = 4, fig.width = 6}

# Hematology deep dive
fda_approvals_long %>% 
  filter(Therapeutic_Area == "Hematology") %>% 
  select(Brand_Name, Demographic, Percentage, Therapeutic_Area, Disease, Approval_Year) %>% 
  filter(Demographic == "White" | Demographic == "Black" | Demographic == "Asian" | Demographic == "Hispanic") %>%
  filter(Percentage > -1) %>% 
  unique() %>% 
  ggplot(aes(Demographic, Percentage)) + 
  geom_jitter(width = 0.2, aes(colour=Demographic)) + 
  scale_y_continuous(breaks = round(seq(min(0), max(100), by = 10),1)) +
  facet_wrap(~Disease) + 
  theme(legend.position = "top", legend.title = element_blank()) +
  ggtitle("Diversity in Hematology FDA approvals [2015-19]")

```


```{r, fig.height = 6, fig.width = 10}
# Cardiovascular Diseases deep dive
fda_approvals_long %>% 
  filter(Therapeutic_Area == "Cardiovascular Diseases") %>% 
  select(Brand_Name, Demographic, Percentage, Therapeutic_Area, Disease, Approval_Year) %>% 
  filter(Demographic == "White" | Demographic == "Black" | Demographic == "Asian" | Demographic == "Hispanic") %>%
  filter(Percentage > -1) %>% 
  unique() %>% 
  ggplot(aes(Demographic, Percentage)) + 
  geom_jitter(width = 0.2, aes(colour=Demographic)) + 
  scale_y_continuous(breaks = round(seq(min(0), max(100), by = 10),1)) +
  facet_wrap(~Disease) + 
  theme(legend.position = "top", legend.title = element_blank()) +
  ggtitle("Diversity in Cardiovascular Diseases FDA approvals [2015-19]")
```


```{r, fig.height = 6, fig.width = 10}
# Psychiatry Diseases deep dive
fda_approvals_long %>% 
  filter(Therapeutic_Area == "Psychiatry") %>% 
  select(Brand_Name, Demographic, Percentage, Therapeutic_Area, Disease, Approval_Year) %>% 
  filter(Demographic == "White" | Demographic == "Black" | Demographic == "Asian" | Demographic == "Hispanic") %>%
  filter(Percentage > -1) %>% 
  unique() %>% 
  ggplot(aes(Demographic, Percentage)) + 
  geom_jitter(width = 0.2, aes(colour=Demographic)) + 
  scale_y_continuous(breaks = round(seq(min(0), max(100), by = 10),1)) +
  facet_wrap(~Disease) + 
  theme(legend.position = "top", legend.title = element_blank()) +
  ggtitle("Diversity in Psychiatry Diseases FDA approvals [2015-19]")
```


```{r, fig.height = 4, fig.width = 8}

# Endocrinology and Metabolism Diseases deep dive
fda_approvals_long %>% 
  filter(Therapeutic_Area == "Endocrinology and Metabolism") %>% 
  select(Brand_Name, Demographic, Percentage, Therapeutic_Area, Disease, Approval_Year) %>% 
  filter(Demographic == "White" | Demographic == "Black" | Demographic == "Asian" | Demographic == "Hispanic") %>%
  filter(Percentage > -1) %>% 
  unique() %>% 
  ggplot(aes(Demographic, Percentage)) + 
  geom_jitter(width = 0.2, aes(colour=Demographic)) + 
  scale_y_continuous(breaks = round(seq(min(0), max(100), by = 10),1)) +
  facet_wrap(~Disease) + 
  theme(legend.position = "top", legend.title = element_blank()) +
  ggtitle("Diversity in Endocrinology and Metabolism FDA approvals [2015-19]")


```





```{r, include = FALSE}

TA_race_graph <- fda_approvals_long %>% 
  select(Brand_Name, Demographic, Percentage, Therapeutic_Area, Approval_Year) %>% 
  filter(Demographic == "White" | Demographic == "Black" | Demographic == "Asian" | Demographic == "Other") %>%
  filter(Percentage > -1) %>% 
  unique() %>% 
  ggplot(aes(Demographic, Percentage)) + 
  geom_jitter(width = 0.2, aes(colour=Demographic)) + 
  facet_wrap(~Therapeutic_Area) + 
  ggtitle("a.i Race participation by Therapeutic Area in clinical trials for FDA approvals [2015-19]")

TA_race_trend_graph <- fda_approvals_long %>% 
  select(Brand_Name, Demographic, Percentage, Therapeutic_Area, Approval_Year) %>% 
  filter(Demographic == "White" | Demographic == "Black" | Demographic == "Asian" | Demographic == "Other") %>%
  filter(Percentage > -1) %>% 
  unique() %>% 
  ggplot(aes(Demographic, Percentage)) + 
  geom_jitter(width = 0.2, aes(colour=Approval_Year)) + 
  facet_wrap(~Therapeutic_Area) + 
  ggtitle("a.ii. Race participation by Therapeutic Area & Year in clinical trials for FDA approvals [2015-19]")

TA_ethnicity_graph <- fda_approvals_long %>% 
  select(Brand_Name, Demographic, Percentage, Therapeutic_Area, Approval_Year) %>% 
  filter(Demographic == "Hispanic" | Demographic == "Non_Hispanic") %>%
  filter(Percentage > -1) %>% 
  unique() %>% 
  ggplot(aes(Demographic, Percentage)) + 
  geom_jitter(width = 0.2, aes(colour=Demographic)) + 
  facet_wrap(~Therapeutic_Area) + 
  ggtitle("b.i Ethnicity participation by Therapeutic Area in clinical trials for FDA approvals [2017-19]")

TA_ethnicity_trend_graph <- fda_approvals_long %>% 
  select(Brand_Name, Demographic, Percentage, Therapeutic_Area, Approval_Year) %>% 
  filter(Demographic == "Hispanic" | Demographic == "Non_Hispanic") %>%
  filter(Percentage > -1) %>% 
  unique() %>% 
  ggplot(aes(Demographic, Percentage)) + 
  geom_jitter(width = 0.2, aes(colour=Approval_Year)) + 
  facet_wrap(~Therapeutic_Area) + 
  ggtitle("b.ii. Ethnicity participation by Therapeutic Area & Year in clinical trials for FDA approvals [2017-19]")

```


```{r, fig.height = 10, fig.width = 10}
TA_race_graph
```


```{r, fig.height = 10, fig.width = 10}
TA_race_trend_graph
```


```{r, fig.height = 10, fig.width = 10}
TA_ethnicity_graph
```


```{r, fig.height = 10, fig.width = 10}
TA_ethnicity_trend_graph
```

# 4) Ability to assess efficacy/safety differences

```{r}

brks <- c(0, 0.25, 0.5, 0.75, 1)

Approvals_18_19 <- fda_approvals %>% 
  filter(Approval_Year >= 2018) %>% 
  group_by(Approval_Year) %>% 
  summarize(Count = n())
  
total_18_19 <- sum(Approvals_18_19$Count)

fda_approvals_long %>% 
  filter(Compared == "yes") %>% 
  filter(Approval_Year >= 2018) %>%
  group_by(Comparison) %>% 
  summarize(Count = round(100 * n() / (total_18_19*10), 0)) # Divide by 10 bc 10 variables for demographic 

fda_approvals_long %>% 
  filter(Compared == "yes") %>% 
  filter(Approval_Year >= 2018) %>%
  group_by(Comparison) %>% 
  summarize(Count = round(100 * n() / (total_18_19*10), 0)) %>% # Divide by 10 bc 10 variables for demographic 
  ggplot() + 
  geom_col(aes(Comparison, Count)) + 
  #geom_text(aes(label=Count), position = position_dodge(0.9), vjust=-0.3, size=3) +
  xlab("Comparison Type") +
  ylab("Percent of Approvals") + 
  ggtitle("Ability to assess efficacy/safety in trial data for FDA approval (2018,19)")

#fda_approvals_long %>% 
#  filter(Approval_Year >= 2018) %>%
#  #mutate(Perc = Count / total_18_19) %>% 
#  ggplot() +
#  geom_bar(aes(Comparison, fill = Compared)) + 
#  scale_y_continuous(labels = scales::percent(brks)) + 
#  ggtitle("Percent of studies that were able to assess efficacy/safety across Age, Race, Sex [FDA approvals 2018,19]")

```


